---
title: "ggplot2 lab"
author: "Lab assignment authored by Joanna Lankester"
subtitle: Lab completed by Joshua Okero
output:
  html_document:
    df_print: paged
---

# Instructions
Go through this R notebook sequentially. Questions are numbered with the letter Q. 
Some questions just have code you need to run. Questions requiring code or a response will have a *. 
Please remember to **title** plots and **label** axes (including units, where needed).

# Setup
First we'll load libraries and data needed for this lab.

### Q1
Run this code first. 
```{r}
library(tidyverse)
library(reshape2)
# in billions of dollars
df <- melt(USPersonalExpenditure,varnames=c("category","year"),value.name = "amount")

mpg %>% 
    filter(manufacturer=="ford" | manufacturer=="toyota" | manufacturer=="nissan") %>%  
    group_by(manufacturer,class) %>% 
    tally() %>% 
    group_by(manufacturer) %>% 
    mutate(total = sum(n)) %>% 
    group_by(manufacturer) %>% 
    mutate(percent_of_brand = 100*n/total) %>% 
    select(manufacturer,class,n,percent_of_brand) -> df_car
```

# Time series plots

### Q2

Run this command to understand what the dataset looks like. 
```{r}
df %>% head(10)
```
### Q3* 
Here are two ways of making bar plots of the amount people spent each year in the U.S. on various categories (in billions of dollars). 
Fill in the title (ending with your initials) and label. 
```{r}
ggplot(df,aes(x=year,y=amount,fill=category)) + 
    geom_col() +
    labs(title="Amount of Money US Consumers Spent Each Year",y="label (including units)")

ggplot(df,aes(x=year,y=amount,fill=category)) + 
    geom_col(position="dodge") +
    labs(title="Amount of Money US Consumers Spent Each Year",y="label (including units)")
```

### Q4*
However, there is usually a better way to represent time data.
Copy and paste one of the blocks of code from above and change it to the type of plot that is generally used for time series data.

Hint: change will include a geom change, and another small item
```{r}
ggplot(df,aes(x=year,y=amount,fill=category, color = category)) + 
    geom_line() +
    labs(title="Amount of Money US Consumers Spent Each Year",y="label (including units)")

```

### Q5 OPTIONAL
Due to the difference in order of magnitude for the expenditures, it's difficult to see how rapidly some of them grew. 
Add a scale change component to make it easier to see.
```{r}

```
# Plotting a histogram, bar plot, and box plot

### Q6
Run this command to see what the dataset looks like.
```{r}
tips %>% head()

```
### Q7*
The geom_histogram() requires either a number of bins or a binwidth. Choose one that makes the plot look nice. Add title and label appropriately.
```{r}
ggplot(tips,aes(x=total_bill)) + 
    geom_histogram(bin_width=1)+labs(title = "Total Bills Amounts A Waiter Received for Several Months") 
```
### Q8*
Now modify the histogram to become a boxplot instead. Hint: this should just involve a geom change. 
Also, change the "x=" to "y=". Update labels as needed.

Note: this is an intermediate step (on the way to Q9) that will make a plot that doesn't yet
give much information. 
```{r}
ggplot(tips,aes(y=total_bill)) + 
    geom_boxplot()+labs(title = "Total Bills Amounts A Waiter Received for Several Months") + theme(axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank())
```
### Q9*
Modify to make a boxplot by day of the week. 
Hint: you will need an additional specification in the "aes(...)" section.

```{r}
ggplot(tips,aes(x= day, y=total_bill))+ 
    geom_boxplot()+labs(title = "Total Bills Amounts A Waiter Received for Certain Days",x="Day of the week")
```
### Q10* (response)
Suppose we want to create a bar plot of the **average** bill for dining by day of the week. 
You have a rough idea of how expensive the restaurant bills are based on your histogram
and box plots above. What do you think the following plot might be showing? (i.e. is it showing
average bill like we want, or something else, and if something else, what is it?)

*Replace this text with your answer*

```{r}
#The plot shows the sum of all bills for a respetive day
```
### Q11*
Here is the way around it. For future labs, we will learn how to do pre-processing of the 
data so that it is easier to make plots without this syntax.  

- Customize it by flipping the coordinate to make it a horizontal plot. Do this without changing the mapping of x= and y=.  
- Add a fill="colorname" option inside the stat_summary parentheses, filling in colorname whatever color you'd like. Here is a list of options: (see the assignment on Canvas)
- Title/label as needed.

```{r}
ggplot(tips,aes(x=day)) + 
    stat_summary(aes(y=total_bill),fill="chartreuse3",fun="median",geom = "bar")+labs(title="Number of Tips A Waiter Received on Certain Days")+coord_flip()
```
# Plotting proportions

### Q12
Run this command to see what the following dataset looks like: 
```{r}
df_car %>% head()
```
### Q13
Here's a baseline plot to start with. We can add different geoms to produce plots. 
Run the following line of code. Note that no plot will be generated. Instead, we 
will obtain an object that we can add to in order to create plots.

```{r}
car_plot <- ggplot(df_car,aes(x=manufacturer,y=percent_of_brand,fill=class)) 
```

### Q14*
Add a geom to car_plot to make a stacked bar plot. (What else is needed to complete the plot?)
```{r}
# Hint: 
car_plot + geom_bar(position="fill", stat="identity")+labs( title = "Car Class Percentage by Brand")

```

### Q15*
Convert the stacked bar plot to a grouped bar chart. 
```{r}
ggplot(df_car,aes(x=manufacturer,y= percent_of_brand, fill=class))+geom_bar(position="dodge", stat="identity")+labs( title = "Percentage of a Car Class by Brand")
```


### Q16* (response only)

Here is the code for a pie chart. Answer these questions (no coding required): 

a. Why is this pie chart made with a geom_col()? In other words, what is the difference between a stacked bar and a pie chart, from the Grammar of Graphics perspective? 

*Type your response here*
A pie chart makes it easy to compare proportions. On the other hand a stacked bar plot represents absolute values of variables.

In this case,a pie chart is used because easily comparing proportions is preferred. 

b. What is the line "facet_grid" doing? If you're not sure, add this line on to your grouped bar 
plot (it won't be a correct plot, but you'll see the change that is made)

*Type your response here*
facet_grid() makes a grid of plots that show different subsets of the data.

```{r}
ggplot(df_car,aes(x="",y=percent_of_brand,fill=class)) + 
    geom_col() + 
    coord_polar("y") +
    theme_void() + 
    facet_grid(~manufacturer)+  labs(title= "Proportion of A Car Class by Brand")
```

### Q17*
Finally, choose 3 built-in themes that you'd like to try. 
Choose 3 plots from this lab and paste them here, and apply a theme to each. Make these different from the default theme.

```{r}
# theme 1: 
ggplot(df_car,aes(x="",y=percent_of_brand,fill=class)) + 
    geom_col() + 
    coord_polar("y") +
    theme_void() + 
    facet_grid(~manufacturer) +theme_classic()+ labs(title= "Proportion of A Car Class by Brand")

```

```{r}
# theme 2:
ggplot(tips,aes(x=day)) + stat_summary(aes(y=total_bill),fill="chartreuse3",fun="median",geom = "bar")+coord_flip() +theme_light()+labs(title="Number of Tips A Waiter Received on Certain Days")

```

```{r}
# theme 3:
car_plot + geom_bar(position="dodge", stat="identity") + theme_linedraw()+labs( title = "Percentage of a Car Class by Brand")
```

# Turning in the lab

### Important: 
- Don't forget to title and label all plots
- You may want to re-run everything from the top if you think you may have changed the code 
without running it at any point. (One easy way is to click at the top, use the "Run"
dropdown, and select "Run All Chunks Below")

### To turn in - 2 files: 
- The .Rmd file that is the R notebook
- The .html file that is produced when you click the "Preview" button